---
title: "wetlandmapR Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{wetlandmapR Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---




```{r message=FALSE, warning=FALSE, include=FALSE}
library(wetlandmapR)
library(rgeos)
library(rgdal)
library(tidyverse)
library(bcmaps)
library(raster)
library(sf)
library(doParallel)
```

The following code demonstrates the use of the [wetlandmapR package](https://github.com/bcgov/wetlandmapR) to classify wetlands from terrain derivatives and external inputs such as satellite imagery using example data provided in the package.  


## Raster data set-up ... 

The following lines of code define an output directory, load in an example digital elevation model (DEM) from the [wetlandmapR package](https://github.com/bcgov/wetlandmapR), and calls the ***wetlandmapR::create_dem_products*** function to generate terrain derivatives using [SAGA-GIS](http://www.saga-gis.org/) and [RSAGA](https://CRAN.R-project.org/package=RSAGA). These terrain derivatives will be used to predict the wetland classes. All of the terrain outputs including an image of Sentinel-2 band 4 are converted to a raster stack by calling the ***wetlandmapR::stack_raster*** function. The resulting raster stack is plotted below. As a pre-processing step, ***wetlandmapR::create_dem_products*** first hydrologically conditions the DEM by filling any sinks, a binary sinks layer is output under the name *SINKS.sgrd*.

```{r fig.width=6, fig.height=6, message=FALSE, warning=FALSE, results = FALSE, fig.cap="Raster stack of SAGA terrain derivatives and Sentinel-2 band 4 image."}
# Set a random seed 
set.seed(42)

#Specify a output directory, we'll use the temporary directory for this example
out_dir<-paste(tempdir(),"/wetlandmapR_expl",sep="")
dir.create(out_dir)

#Load a example digital elevation model DEM from wetlandmapR package 
target_dem <- system.file('extdata','DEM.tif',package = 'wetlandmapR')

#Create SAGA elevation derivatives using the wetlandmapR 'create_dem_products' function
create_dem_products(dem = target_dem, outdir = out_dir)

#Get list of SAGA terrain derivatives within the output directory 
raster_list <- list.files(out_dir, "sdat$", full.names = TRUE)

#Drop the DEM w/ sinks from list
raster_list<-raster_list[!raster_list %in% file.path(out_dir,'ELEV_NoSink.sdat')]

#E.g., Add external Sentinel-2 Band 4 data to the list of raster inputs from wetlandmapR package 
raster_list<-append(raster_list,system.file("extdata", "B4_SENT2.tif", package = "wetlandmapR"))

#Stack all raster inputs using the wetlandmapR 'stack_raster' function 
raster_stack <- stack_rasters(rasters = raster_list, target_raster = target_dem, outdir = out_dir,rastLUTfn = file.path(out_dir,'rastLUT.csv'), aligned=FALSE)

#Plot the new stack of raster objects (co-variate layers)
plot(raster_stack)

```



## Attribute raster values to training points ...

The following code reads in example training points from the [wetlandmapR](https://github.com/bcgov/wetlandmapR) package as a *sf* object. In addition, biogeoclimatic zones (BEC) are loaded internally using the [bcmaps](https://github.com/bcgov/bcmaps) package. Using the raster stack plotted above as input, the ***wetlandmapR::grid_values_at_sp*** function attributes the raster values intersecting the training points grouped by the AOI polygons, in this case BEC zones. The results are written to the output directory as a CSV.


```{r message=FALSE, warning=FALSE, results=FALSE}
#Load training data as sp  
training_points<-st_read(system.file('extdata', 'TrainingPoints.shp', package = 'wetlandmapR'))

#Load BEC zones from bcmaps as AOIs
aoi_polys<-bec(ask=F) %>% st_crop(training_points)

#Convert aoi_polys to sp
aoi_polys<-as_Spatial(aoi_polys)


#Assign URL to output CSV
attributed_csv <- file.path(out_dir,"train_pnts_attributed.csv")

#Attribute training points with values at intersection with co-variate layers
grid_values_at_sp(x=raster_stack,
                  y=training_points,
                  filename = attributed_csv,
                  aoi = aoi_polys)


```

##  Setup predictor list and raster LUT...

Read in the *rastLUT.csv* file which was output by ***wetlandmapR::stack_rasters***, the first column of this CSV contains the URL to each co-variate raster layer, while the second column indicates the name of the layer. Here we get a list of the raster layer names from the *rastLUT* table, and create a new directory for our model outputs. 

```{r message=FALSE, warning=FALSE}
#Read in rastLUT table as data.frame 
rastLUT <- read.csv(file.path(out_dir,'rastLUT.csv'),
                    header = FALSE,
                    stringsAsFactors = FALSE)

#View rastLUT table 
rastLUT

#Get the list of co-variate predictor names (2nd column)
predList <- rastLUT[,2]


#Create an output directory within 'out_dir'
dir.create(file.path(out_dir,"output"))

```


## Run model and diagnostics...

Below we build a random forest wetland classification model by calling the ***wetlandmapR::wetland_model*** function using the training data that we attributed with the co-variate raster layers above. We declare the response field to be the *T_W_Class* field. We specify that we want to use the *BGC_ZONE* column to identify separate AOIs. We also define the *SINKS* column to be a factor variable. The results include a is a list containing *model.build* objects output from the [ModelMap](https://cran.r-project.org/web/packages/ModelMap/index.html) package for each AOI. 

```{r message=FALSE, warning=FALSE}
#Fit a random forest model 
model.out <- wetland_model(qdatafn = attributed_csv,
                           model.type = "RF",
                           model.folder = file.path(out_dir,"output"),
                           unique.rowname = "OBJECTID",
                           predList = predList,
                           predFactor = c('SINKS'),
                           response.name = "T_W_Class",
                           response.type = "categorical",
                           seed = 44,
                           response.target = as.vector(unique(training_points$T_W_Class)),
                           aoi.col = "BGC_ZONE")
```


## Create map(s) from model...

Finally we call the ***wetlandmapR::wetland_map*** function which generates raster prediction surfaces using the model outputs from ***wetlandmapR::wetland_model***, i.e., *model.out*. We point the function to the output directory assigned to *wetland_model* above. We also pass the *rastLUT* data frame, and and BEC AOI SpatialPolygons, the AOI column is defined as the *ZONE* column. 

```{r warning=FALSE, message=FALSE, results=FALSE}
#Map random forest results 
wetland_map(model.out = model.out,
            model.folder = file.path(out_dir,"output"),
            rastLUTfn = rastLUT,
            aoi = aoi_polys,
            aoi.col = "ZONE")
```


## Attribute upstream basin stats...

Optionally, after mapping has been completed, upstream basin attributes can be attributed to any point within the study area using ***wetlandmapR::upstream_basin_stats***.

### Set up the GRASS-GIS environment:

First we set up a [**GRASS-GIS**](https://grass.osgeo.org/) environment using the [**rgrass7**](https://cran.r-project.org/web/packages/rgrass7/index.html) package. This is done by calling ***wetlandmapR::set_grass_env***. This function requires that the path the the [**GRASS-GIS**](https://grass.osgeo.org/) binaries be provided, in addition to a list of *raster* objects to be summarized, such as those output above, and associated vector of raster layer names. As this function relies on the  [**GRASS-GIS**](https://grass.osgeo.org/) *r.stream.extract* module to derive a stream network a accumulation threshold in pixels must be defined. 

```{r warning=FALSE, message=FALSE, results=FALSE}
#!!Change for your current GRASS-GIS installation!! 
gisbase<-'/usr/lib/grass76/'

#Get list of raster layers and names 
lyr_lst<-list()
lyr_names<-c()
for(lyr in  c(1:dim(raster_stack)[3]))
{
  lyr_lst[[lyr]]<-raster(rastLUT[lyr,1])
  lyr_names[lyr]<-rastLUT[lyr,2]
}

#Write the raster objects to a GRASS-GIS environment, derive streams and create GRASS DEM derivatives 'r.watershed' 
set_grass_env(gisbase=gisbase,
              DEM=raster(target_dem),
              lyr_list=lyr_lst,
              lyr_names=lyr_names,
              acc_thresh = 1000)
```


### Calculate upstream basin statistics 

Below we convert the DEM loaded above into a data frame with the *X* and *Y* coordinates of each pixel, and attribute a unique identifier field *UID*. We take a sample of these pixels to serve as pour points for the purpose of this example. We then compute the mean and standard deviation of upstream elevation and topographic wetness for each pour point using the ***wetlandmapR::run_basin_stats*** function. 

```{r message=FALSE, warning=FALSE}
#Create a data frame of pour points for purpose of example 
pour_pnts<-as.data.frame(raster(target_dem),xy=T)
pour_pnts<-pour_pnts[complete.cases(pour_pnts),]
colnames(pour_pnts)<-c('X','Y','UID')
pour_pnts$UID<-c(1:nrow(pour_pnts))

#Take a sample, processing can take a long time when n is large 
pour_pnts<-pour_pnts[sample(c(1:nrow(pour_pnts)),50),]

#View pour point table 
head(pour_pnts)

#Set to number of cores minus 1 
cores<-parallel::detectCores()-1

#For the random pour points calculate upstream basin mean and standard deviation for elevation and topographic wetness.
basin_stats<-run_basin_stats(pour_pnts = pour_pnts,
                             covar_rast = c('ELEV','TOPOWET','ELEV','TOPOWET'),
                             stat_vect = c('MEAN','MEAN','STDDEV','STDDEV'),
                             procs = cores)

#View upstream basin statistics 
head(basin_stats)
```

