---
title: "wetlandmapR Example"
output: rmarkdown::html_vignette
author: "Deepa Filatow, Gillian Harvey, Hunter Gleason"
vignette: > 
  %\VignetteIndexEntry{wetlandmapR Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---
  
  
The following code demonstrates the use of the [wetlandmapR](https://github.com/bcgov/wetlandmapR) package to classify wetlands using terrain derivatives and external inputs such as satellite imagery using example data provided in the package.
  
## Using the wetlandmapr_dokr ...
  
A docker image based on the [**rocker/geospatial:3.6.3**](https://hub.docker.com/r/rocker/geospatial) image has been assembled for running the [wetlandmapR package](https://github.com/bcgov/wetlandmapR) package. Assuming a local installation of Docker, the [huntgdok/wetlandmapr:3.6.3](https://hub.docker.com/repository/docker/huntgdok/geospat) image can be downloaded by passing the command: 
  
```{bash eval=FALSE}
docker pull huntgdok/wetlandmapr:3.6.3
```
once the image is in place, it can be run using the following Docker command
  
```{bash eval=FALSE}
docker run -e PASSWORD=APassword -p 8787:8787 --rm huntgdok/wetlandmapr:3.6.3
```
  
where *APassword* is any password you want. To use with sudo privileges, add *-e ROOT=TRUE* to call. To view the running docker, go to the local URL [localhost](http://localhost:8787) *http://localhost:8787* in your internet browser. When prompted for a username and password, supply *rstudio* as a username, and the password that you set with *docker run*. You should then have a R-Studio session running with all the R and third party dependencies required for running the [**wetlandmapR**](https://github.com/bcgov/wetlandmapR) package. 
  
## Load dependencies ... 
  
```{r message=FALSE, warning=FALSE}
library(wetlandmapR)
library(rgeos)
library(rgdal)
library(tidyverse)
library(bcmaps)
library(raster)
library(sf)
library(doParallel)
```


## Raster data set-up ... 

The following lines of code define an output directory, load in an example digital elevation model (DEM) from the [wetlandmapR package](https://github.com/bcgov/wetlandmapR), and calls the ***wetlandmapR::create_dem_products*** function to generate terrain derivatives using [SAGA-GIS](http://www.saga-gis.org/) and [RSAGA](https://CRAN.R-project.org/package=RSAGA). These terrain derivatives will be used to predict the wetland classes. All of the terrain outputs including an image of Sentinel-2 band 4 are converted to a raster stack by calling the ***wetlandmapR::stack_raster*** function. The resulting raster stack is plotted below. As a pre-processing step, ***wetlandmapR::create_dem_products*** first hydrologically conditions the DEM by filling any sinks, a binary sinks layer is output under the name *SINKS.sgrd*.

```{r fig.width=6, fig.height=6, message=FALSE, warning=FALSE, results = FALSE, fig.cap="Raster stack of SAGA terrain derivatives and Sentinel-2 band 4 image."}
# Set a random seed 
set.seed(42)

#Specify a output directory, we'll use the temporary directory for this example
out_dir<-paste(tempdir(),"/wetlandmapR_expl",sep="")
dir.create(out_dir)

#Load a example digital elevation model DEM from wetlandmapR package 
target_dem <- system.file('extdata','DEM.tif',package = 'wetlandmapR')

#Load freshwater atlas stream data
streams <- system.file('extdata','streams.shp',package = 'wetlandmapR')

#Create SAGA elevation derivatives using the wetlandmapR 'create_dem_products' function, burn FWA streams into DEM by 5 m ...
create_dem_products(dem=target_dem,stream_vec = streams, burn_val=5, outdir=out_dir)

#Get list of SAGA terrain derivatives within the output directory 
raster_list <- list.files(out_dir, "sdat$", full.names = TRUE)

#Drop the DEM w/ sinks from list
raster_list<-raster_list[!raster_list %in% file.path(out_dir,'ELEV_NoSink.sdat')]

#E.g., Add external Sentinel-2 Band 4 data to the list of raster inputs from wetlandmapR package 
raster_list<-append(raster_list,system.file("extdata", "B4_SENT2.tif", package = "wetlandmapR"))

#Stack all raster inputs using the wetlandmapR 'stack_raster' function 
raster_stack <- stack_rasters(rasters = raster_list, target_raster = target_dem, outdir = out_dir,rastLUTfn = file.path(out_dir,'rastLUT.csv'), aligned=FALSE)

#Plot the new stack of raster objects (co-variate layers)
plot(raster_stack)

```



## Attribute raster values to training points ...

The following code reads in example training points from the [wetlandmapR](https://github.com/bcgov/wetlandmapR) package as a *sf* object. In addition, biogeoclimatic zones (BEC) are loaded internally using the [bcmaps](https://github.com/bcgov/bcmaps) package. Using the raster stack plotted above as input, the ***wetlandmapR::grid_values_at_sp*** function attributes the raster values intersecting the training points grouped by the AOI polygons, in this case BEC zones. The results are written to the output directory as a CSV.


```{r message=FALSE, warning=FALSE, results=FALSE}
#Load training data as sp  
training_points<-st_read(system.file('extdata', 'TrainingPoints.shp', package = 'wetlandmapR'))

#Load BEC zones from bcmaps as AOIs
aoi_polys<-bec(ask=F) %>% st_crop(training_points)

#Plot AOIs and training points
ggplot()+geom_sf(data=aoi_polys,aes(fill=ZONE))+
  geom_sf(data=training_points)

#Convert aoi_polys to sp
aoi_polys<-as_Spatial(aoi_polys)


#Assign URL to output CSV
attributed_csv <- file.path(out_dir,"train_pnts_attributed.csv")

#Attribute training points with values at intersection with co-variate layers
grid_values_at_sp(x=raster_stack,
                  y=training_points,
                  filename = attributed_csv,
                  aoi = aoi_polys)


```

##  Setup predictor list and raster LUT ...

Read in the *rastLUT.csv* file which was output by ***wetlandmapR::stack_rasters***, the first column of this CSV contains the URL to each co-variate raster layer, while the second column indicates the name of the layer. Here we get a list of the raster layer names from the *rastLUT* table, and create a new directory for our model outputs. 

```{r message=FALSE, warning=FALSE}
#Read in rastLUT table as data.frame 
rastLUT <- read.csv(file.path(out_dir,'rastLUT.csv'),
                    header = FALSE,
                    stringsAsFactors = FALSE)

#View rastLUT table 
rastLUT

#Get the list of co-variate predictor names (2nd column)
predList <- rastLUT[,2]


#Create an output directory within 'out_dir'
dir.create(file.path(out_dir,"output"))

```


## Run model and diagnostics ...

Below we build a random forest wetland classification model by calling the ***wetlandmapR::wetland_model*** function using the training data that we attributed with the co-variate raster layers above. We declare the response field to be the *T_W_Class* field. We specify that we want to use the *BGC_ZONE* column to identify separate AOIs. We also define the *SINKS* column to be a factor variable. The results include a is a list containing *model.build* objects output from the [ModelMap](https://cran.r-project.org/web/packages/ModelMap/index.html) package for each AOI. 

```{r message=FALSE, warning=FALSE}
#Fit a random forest model 
model.out <- wetland_model(qdatafn = attributed_csv,
                           model.type = "RF",
                           model.folder = file.path(out_dir,"output"),
                           unique.rowname = "OBJECTID",
                           predList = predList,
                           predFactor = c('SINKS'),
                           response.name = "T_W_Class",
                           response.type = "categorical",
                           seed = 44,
                           response.target = as.vector(unique(training_points$T_W_Class)),
                           aoi.col = "BGC_ZONE")
```


## Create map(s) from model ...

Finally we call the ***wetlandmapR::wetland_map*** function which generates raster prediction surfaces using the model outputs from ***wetlandmapR::wetland_model***, i.e., *model.out*. We point the function to the output directory assigned to *wetland_model* above. We also pass the *rastLUT* data frame, and BEC AOI SpatialPolygons, the AOI column is defined as the *ZONE* column. 

```{r warning=FALSE, message=FALSE, results=FALSE}
#Map random forest results 
wetland_map(model.out = model.out,
            model.folder = file.path(out_dir,"output"),
            rastLUTfn = rastLUT,
            aoi = aoi_polys,
            aoi.col = "ZONE")
```


## Plot map results

Plot the output classified ERDAS images by AOI ...

```{r warning=FALSE, message=FALSE}
#Get .img file paths from temporary directory 
SBS.dir<-list.files(paste(tempdir(),'/wetlandmapR_expl/output/',sep=''),pattern = "SBS")
ESSF.dir<-list.files(paste(tempdir(),'/wetlandmapR_expl/output/',sep=''),pattern = "ESSF")
SBS.img<-paste(SBS.dir,'_map.img',sep='')
ESSF.img<-paste(ESSF.dir,'_map.img',sep='')

#Plot map results for each AOI
plot(as.factor(raster(paste(tempdir(),'/wetlandmapR_expl/output/',SBS.dir,'/',SBS.img,sep=''))))
plot(as.factor(raster(paste(tempdir(),'/wetlandmapR_expl/output/',ESSF.dir,'/',ESSF.img,sep=''))))
```

